Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-08-24 ebayes_rnaseq_heterosis.tex
update alternative methods to include Ji

Existing \RNAseq{} data analysis methods are not specifically designed to identify gene expression heterosis, but we adapted the approaches in \edgeR{} and {\tt baySeq} to provide a measure of the strength of heterosis for each gene such that large values of this measure indicate support for heterosis.

@@

{\tt baySeq} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct. In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene. We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$. Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for overdominance.

>>>>

We compared our method to that of \cite{ji2014estimation}, which assumes normality in the response, by modeling the logarithm of the RNAseq count plus one divided by the normalization factor, i.e. $\log([Y_{gvi}+1]/\gamma_{vi})$. Use of the normalization factor here provides two advantages: 1) counts are properly scaled by the thoroughness of the sequencing of the sample and 2) for genes with no variability within variety (which actually occurs in our maize data set), use of the normalization factors allows the approach of \citeauthor{ji2014estimation} to still execute.

In addition to the approach of \citeauthor{ji2014estimation}, we modified two existing \RNAseq{} approaches, \edgeR{} and \baqSeq{}, for use in the overdominance context. For each method, we attempted to provide a measure of the strength of overdominance for each gene such that large values of this measure indicate support for heterosis.

@@

\baqSeq{} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct. In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene. We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$. Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for overdominance.
